Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 paper.tex
ADdress a few points

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes. \revpoint{2}{5}
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
osmond_estimating_2021,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches \revpoint{2}{6}
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020}
has been the most widely used and studied,
and until recently consistently out-performed other methods in terms
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has successfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}.
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more
scalable~\citep{deng_robust_2024}.
The second class of inference methods that has been of recent interest
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023}
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different
principles, but share some common properties.
Firstly, they are all
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model. \revpoint{2}{7}
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when
producing the initial ARG. General purpose ARG dating methods are
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the
purposes of this paper) the ARGs that heuristic methods produce
lack some of the topological information present in the output of ARGweaver or SINGER.
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.

>>>>

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in genetics~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes. \revpoint{2}{5}
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
osmond_estimating_2021,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are probabilistic approaches \revpoint{2}{6}
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020}
has been the most widely used and studied,
and until recently consistently out-performed other methods in terms
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has successfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}.
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more
scalable~\citep{deng_robust_2024}.
The second class of inference methods that has been of recent interest
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023}
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different
principles, but share some common properties.
Firstly, they are all
heuristic and approximation driven, favouring computational
scalability over a basis in a probabilistic model. \revpoint{2}{7}
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when
producing the initial ARG. General purpose ARG dating methods are
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the
purposes of this paper) the ARGs that heuristic methods produce
lack some of the topological information present in the output of ARGweaver or SINGER.
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.
